---
title: "import data from fj workspaces"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{import data from fj workspaces}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = F}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(knitr.kable.NA = '', width = 80)
library(flowWorkspace)
library(CytoML)
```

Install necessary packages in RStudio:
```{r setup, eval=F,message=F}
install.packages("devtools")
devtools::install_github("Close-your-eyes/fcexpr")
install.packages("tidyverse")
library(fcexpr)
```
```{r setup2, message=F}
library(tidyverse)
```

Many information are stored in flowjos wsp file which actually is an xml file. This includes gated event counts, statistics like MFI and meta data from the FCS files like keywords. The wsp file can be accessed without a flowjo dongle - you can open the it with in a text editor (e.g. [BB Edit](https://www.barebones.com/products/bbedit/index.html) on a Mac) or read it with the XML2 package in R.   
Based on the latter I wrote functions that extract relevant data from a flowjo 10 wsp (more precise 10.7.1). Let me know of conflicts with other versions. It is very simple and works as follows. After you gated your populations, save the wsp and run:

```{r,include=F}
dir <- base::system.file("extdata", package = "fcexpr")
utils::untar(base::system.file("extdata", "Part_3.tgz", package = "fcexpr"), exdir = dir)
wd <- file.path(dir, "Part_3")
ws <- list.files(file.path(wd, "FJ_workspace"), full.names = T)
sd <- openxlsx::read.xlsx(file.path(dir, "Part_3", "sampledescription.xlsx"))
```

```{r}
ps <- fcexpr::wsx_get_popstats(ws)
```

ws is the file path (url) to the wsp on your disk (e.g. "/Users/vonskopnik/Documents/07-Oct-2021.wsp"). On a Mac you can copy this path with CMD+ALT+C. The prefix (fcexpr::) indicates the package a function comes from. This is important to avoid ambiguities in case there are two functions with equal names from different packages.  
If there are calculated statistics we get a list of length 2. Index 1 hold a table (data.frame) with gated event counts, index 2 a table with statistics.  
The column names are very descriptive so I will not explain them, except for one: eventsInside indicates if the events inside (1) or outside (0) were relevant for subsequent gates. Please check the content of those data frames by yourself.
```{r}
length(ps)
names(ps)

# columns names
names(ps[[1]])
names(ps[[2]])

ps_counts <- ps[["counts"]]
ps_stats <- ps[[2]]
```

Most likely additional information (meta data) are necessary to plot the data or analyze them statistically. This may be achieved by joining a sampledescription to ps_counts or ps_stats. You may create such table manually or have it created in a standardized way with **fcexpr::sync_sampldescription** (see another vignette for the whole concept). Alternatively, your wsp may hold annotations in the form of keywords. These can be extracted as well:

```{r}
keys <- fcexpr::wsx_get_keywords(ws)
```

This returns a list of data frames. One index for every FCS file in the wsp. In this example there are 48 FCS files. The data.frame at index 1 (first FCS file) has 211 rows, one for each keyword, and 2 columns, providing the name of the keyword and its value.
```{r}
length(keys)
nrow(keys[[1]])
names(keys[[1]])
```

To join the keywords to ps_counts or ps_stats we have to make it a data.frame first.
```{r}
# make it a data.frame
keys_df <- do.call(rbind, keys) # base R
keys_df <- dplyr::bind_rows(keys) # dplyr-style
# create a FileName-column
keys_df$FileName <- rep(names(keys), sapply(keys,nrow))
# select relevant keywords
# example: Cytometer name ($CYT) and Operator ($OP)
keys_df <- keys_df[which(keys_df$name %in% c("$CYT", "$OP")),] # base R
keys_df <- dplyr::filter(keys_df, name %in% c("$CYT", "$OP")) # dplyr-style
# make it a wider data.frame (pivot_wider and pivot_longer are a bit difficult to get your head around but are very powerful)
keys_df <- tidyr::pivot_wider(keys_df, names_from = name, values_from = value)
nrow(keys_df)
names(keys_df)
```

Now we have one data.frame of two keywords for every FCS file. Join the meta data to phenotype data. CYT and OP are just two examples but are not very informative. Hence, we also join a sampledescription with more relevant meta data
```{r}
ps_counts2 <- merge(ps_counts, keys_df) # base R
ps_counts2 <- dplyr::left_join(ps_counts, keys_df) # dplyr-style
# sd is a sampledescription data.frame
names(sd)
ps_counts2 <- dplyr::left_join(ps_counts2, sd)
```

ps_counts2 which now holds meta data and phenotype data may be saved to excel for further processing.
```{r, eval=F}
openxlsx::write.xlsx(ps_counts2, "/Users/vonskopnik/Documents/7-Oct-2021.xlsx")
```

It is preferable though to tidy up, filter, select and plot in R. This can be done with base R or with functions from the [tidyverse](https://www.tidyverse.org). The possible operations and sometimes necessary steps to process a data frame for plotting are too exhaustive to cover them here. Also, it requires a little bit of practice. For a number of cases there are a few standard procedures that you may learn. A few arbitrary operations and plots are shown below.

```{r,fig.width=8}
# filter for the GFP+ population only
unique(ps_counts2$Population)
ps_counts2_plot <-
  ps_counts2 %>%
  dplyr::filter(Population == "GFP+")
unique(ps_counts2_plot$Population)
# plot the frequency of GFP+ events with respect to the parent gate
ggplot(ps_counts2_plot, aes(x = incubation.time.h, y = FractionOfParent)) +
  geom_jitter(width = 0.1) +
  theme_bw() +
  ylab("% GFP+ in DAPI-")
```

The incubation time of viral supernatant (x) seems to play a role for the transduction rate (y). But there are more factors which may visualized in one plot.

```{r, include=F}
ps_counts2_plot$incubation.time.h <- factor(ps_counts2_plot$incubation.time.h, levels = sort(as.numeric(unique(ps_counts2_plot$incubation.time.h))))
ps_counts2_plot$EGF.ng.ml <- factor(ps_counts2_plot$EGF.ng.ml, levels = sort(as.numeric(unique(ps_counts2_plot$EGF.ng.ml))))
ps_counts2_plot$seeding.density.dilution <- factor(ps_counts2_plot$seeding.density.dilution, levels = sort(as.numeric(unique(ps_counts2_plot$seeding.density.dilution))))
ps_counts2_plot$viral.sup.vol.µl <- factor(ps_counts2_plot$viral.sup.vol.µl, levels = sort(as.numeric(unique(ps_counts2_plot$viral.sup.vol.µl))))
```


```{r,fig.width=8}
ggplot(ps_counts2_plot, aes(x = viral.sup.vol.µl, y = seeding.density.dilution, fill = FractionOfParent)) +
  geom_tile(color = "black") +
  theme_classic() +
  scale_fill_viridis_c() +
  facet_grid(cols = vars(incubation.time.h), rows = vars(FCS.percent))
```


```{r, echo=F}
unlink(wd, recursive = T)
```
