---
title: "synchronizing FCS files with an xlsx file"
date: "2021/10/04"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{synchronizing FCS files with an xlsx file}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = F}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(knitr.kable.NA = '', width = 60)
```

Install necessary packages in RStudio:
```{r setup, eval=F}
install.packages("BiocManager")
BiocManager::install("flowCore")
BiocManager::install("flowWorkspace")
BiocManager::install("CytoML")
install.packages("devtools")
devtools::install_github("Close-your-eyes/fcexpr")
library(fcexpr)
```

Let's create a folder for our latest flow cytometry experiment. Mac-user: To copy the path of a directory use CMD+Alt+C in the Finder. Please don't be puzzled by the strange dir that is shown here. Just use a 'normal' directory on your disk.
```{r}
dir <- base::system.file("extdata", package = "fcexpr")
fcexpr::new_exp(path = dir, name = "my_experiment")
```

A respective folder appears in "path". It is prefixed with todays date as date_prefix = TRUE by default in fcexpr::new_exp (see ?fcexpr::new_exp). It has a default structure and contains a few template folders and files.
```{r}
wd <- file.path(dir, paste0(base::gsub("-", "", base::Sys.Date()), "_my_experiment"))
print(wd)
list.files(wd)
list.files(wd, recursive = T)
```

Next we add FCS files from our first samples in this experiment to FCS_files folder.
```{r, echo=F}
utils::untar(base::system.file("extdata", "Part_1.tgz", package = "fcexpr"), exdir = file.path(wd, "FCS_files"))
```
```{r}
list.files(file.path(wd, "FCS_files", "Part_1"))
```

To initiate an xlsx-file that allows to document meta data of the FCS files we call the sync_sampledescription function. The xlsx-file will always be written to the parent folder of FCS.file.folder.
```{r}
fcexpr::sync_sampledescription(FCS.file.folder = file.path(wd, "FCS_files"), init.columns = c())
```

Now we have the xlsx file, by default named 'sampledescription.xlsx':
```{r}
list.files(file.path(wd))
```

And it looks like this. The FileNames are prefixed with a continuous number.
```{r, echo=F}
knitr::kable(openxlsx::read.xlsx(file.path(wd, "sampledescription.xlsx")))
```

This is now also true for the actual FCS files. **The names of the FCS files and the entries in the xlsx-file have been synchronized.**
```{r}
list.files(file.path(wd, "FCS_files", "Part_1"))
```

When we change the names in the FileName column and run sync_sampledescription synchronization takes place.
```{r, echo=F,include=F}
sd <- openxlsx::read.xlsx(file.path(wd, "sampledescription.xlsx"))
sd[2,1] <- "File_2"
sd[5,1] <- "File_5"
openxlsx::write.xlsx(sd, file.path(wd, "sampledescription.xlsx"), overwrite = T)
```
```{r, echo=F}
knitr::kable(openxlsx::read.xlsx(file.path(wd, "sampledescription.xlsx")))
```
```{r}
fcexpr::sync_sampledescription(FCS.file.folder = file.path(wd, "FCS_files"), write.log = F)
```

The prefixed continuous number is added automatically as well as the .fcs-suffix.
```{r, echo=F}
knitr::kable(openxlsx::read.xlsx(file.path(wd, "sampledescription.xlsx")))
```
```{r}
list.files(file.path(wd, "FCS_files", "Part_1"))
```

Various columns can be added to describe the content (meta data) of the FCS files. **These columns serve to document our experiment and as a side-effect we can use them later on to plot our data.**
```{r, echo=F,include=F}
sd <- openxlsx::read.xlsx(file.path(wd, "sampledescription.xlsx"))
sd$Patient_ID <- sort(rep(1:5, 2))
sd$Staining <- rep(c("FMO", "full"), 5)
sd$Patient_sex <- c(rep("M",4),rep("F",6))
openxlsx::write.xlsx(sd, file.path(wd, "sampledescription.xlsx"), overwrite = T)
```
```{r, echo=F}
knitr::kable(openxlsx::read.xlsx(file.path(wd, "sampledescription.xlsx")))
```

The FCS files have been added in chronological order of acquisition. If we are not happy with that, we can change the order in the xlsx-file and the prefix-number will be adjusted. **NOTE: Never change the content of the identity column and never mix up rows of FileName and identity!!!**

```{r, echo=F,include=F}
sd <- openxlsx::read.xlsx(file.path(wd, "sampledescription.xlsx"))
sd <- sd[c(1,2,9,10,5,6,3,4,7,8),]
openxlsx::write.xlsx(sd, file.path(wd, "sampledescription.xlsx"), overwrite = T)
```
```{r, echo=F}
knitr::kable(openxlsx::read.xlsx(file.path(wd, "sampledescription.xlsx")))
```
```{r}
fcexpr::sync_sampledescription(FCS.file.folder = file.path(wd, "FCS_files"), write.log = F)
```
```{r}
list.files(file.path(wd, "FCS_files", "Part_1"))
```

When we add files of our next samples to the FCS_files folder, the xlsx-file will be appended accordingly by the new files (~synchronized).
```{r, echo=F}
utils::untar(base::system.file("extdata", "Part_2.tgz", package = "fcexpr"), exdir = file.path(wd, "FCS_files"))
```
```{r}
list.files(file.path(wd, "FCS_files"))
fcexpr::sync_sampledescription(FCS.file.folder = file.path(wd, "FCS_files"), write.log = F)
```
```{r, echo=F}
knitr::kable(openxlsx::read.xlsx(file.path(wd, "sampledescription.xlsx")))
```

The function argument exclude.folders in sync_sampledescription specifies folder names that will be ignored for syncing with the xlsx-file. Such folders may contain controls like rainbow beads or compensation controls or similar. In general files which are not be of interest for the statistical analysis of our experiment. If, by accident, these files end up in our xlsx-file since we forgot to copy them to a respective folder we can have them removed by deleting thr FileName, **but only the FileName** from the xlsx-file.

```{r, echo=F,include=F}
sd <- openxlsx::read.xlsx(file.path(wd, "sampledescription.xlsx"))
sd$FileName <- ifelse(grepl("comp", sd$FileName), NA, sd$FileName)
openxlsx::write.xlsx(sd, file.path(wd, "sampledescription.xlsx"), overwrite = T)
```

```{r, echo=F}
knitr::kable(openxlsx::read.xlsx(file.path(wd, "sampledescription.xlsx")))
```
```{r}
fcexpr::sync_sampledescription(FCS.file.folder = file.path(wd, "FCS_files"), write.log = F)
```
```{r, echo=F}
knitr::kable(openxlsx::read.xlsx(file.path(wd, "sampledescription.xlsx")))
```

We can use the meta data columns to paste together the FileNames systematically. Not all information may fit into a FileName though. Hence, additional columns in the xlsx-file allow us to coherently and conveniently document our flow cytometry experiment.
```{r, echo=F,include=F}
sd <- openxlsx::read.xlsx(file.path(wd, "sampledescription.xlsx"))
sd$Patient_ID[11:14] <- sort(rep(6:7, 2))
sd$Staining[11:14] <- rep(c("FMO", "full"), 2)
sd$Patient_sex[11:14] <- c(rep("M",2),rep("F",2))
sd$FileName <- paste0("Pat", sd$Patient_ID, "_", sd$Staining)
openxlsx::write.xlsx(sd, file.path(wd, "sampledescription.xlsx"), overwrite = T)
```
```{r, echo=F}
knitr::kable(openxlsx::read.xlsx(file.path(wd, "sampledescription.xlsx")))
```
```{r}
fcexpr::sync_sampledescription(FCS.file.folder = file.path(wd, "FCS_files"), write.log = F)
```
```{r, echo=F}
knitr::kable(openxlsx::read.xlsx(file.path(wd, "sampledescription.xlsx")))
```
```{r}
list.files(file.path(wd, "FCS_files"), recursive = T)
```
```{r, echo=F}
unlink(wd, recursive = T)
```
