---
title: "synchronizing FCS files with an xlsx file"
date: "2021/10/04"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{synchronizing FCS files with an xlsx file}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = F}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(knitr.kable.NA = '', width = 80)
```

Install necessary packages in RStudio:
```{r setup, eval=F}
install.packages("BiocManager")
BiocManager::install("flowCore")
install.packages("devtools")
devtools::install_github("Close-your-eyes/fcexpr")
library(fcexpr)
```

### Please follow the steps of the vignette and observe what happens on your disk.

## Initiate

Pick a directory on your disk and have a new template folder created using fcexpr::new_exp(). A respective folder appears in "path". It is prefixed with todays date as date_prefix = TRUE by default. It has a default structure and contains a few template folders and files.  
```{r}
dir <- "/Users/vonskopnik/Documents"
fcexpr::new_exp(path = dir, name = "my_experiment")
wd <- file.path(dir, paste0(base::gsub("-", "", base::Sys.Date()), "_my_experiment"))
print(wd)
```

Next, add FCS files to the FCS_files folder. You can use your own files, or use example files that come with this package.
```{r}
# this command uses example files from the fcexpr package
utils::untar(base::system.file("extdata", "Part_1.tgz", package = "fcexpr"), exdir = file.path(wd, "FCS_files"))
```
```{r}
list.files(file.path(wd, "FCS_files", "Part_1"))[1:5]
```

To initiate a sampledescription file call the sync_sampledescription function. The xlsx-file is always written to the parent folder of FCS.file.folder. By default it is named 'sampledescription.xlsx'.
```{r}
fcexpr::sync_sampledescription(FCS.file.folder = file.path(wd, "FCS_files"), init.columns = c())
```
All FCS files have been entered as rows. The FileNames are prefixed with a continuous number to make them unique. The identity column is necessary for the function to identify the FCS unambiguously.  
sampledescription.xlsx:
```{r, echo=F}
knitr::kable(openxlsx::read.xlsx(file.path(wd, "sampledescription.xlsx")))
```

Also the actual FCS files have a prefix now. **The names of the FCS files and the entries in the xlsx-file have been synchronized.**
```{r}
list.files(file.path(wd, "FCS_files", "Part_1"))[1:5]
```


## Add meta data

Additonal columns with meta data can be added. **These columns serve to document the content of FCS files and one can use them group the data when plotting.** They may also be used by another function to hard-code them into the FCS files if this is desired.
```{r, echo=F,include=F}
sd <- openxlsx::read.xlsx(file.path(wd, "sampledescription.xlsx"))
sd$Patient_ID <- sort(rep(1:5, 2))
sd$Staining <- rep(c("FMO", "full"), 5)
sd$Patient_sex <- c(rep("M",4),rep("F",6))
openxlsx::write.xlsx(sd, file.path(wd, "sampledescription.xlsx"), overwrite = T)
```
sampledescription.xlsx:
```{r, echo=F}
knitr::kable(openxlsx::read.xlsx(file.path(wd, "sampledescription.xlsx")))
```

## Rename FCS files

When the names in the FileName column are altered and sync_sampledescription is run synchronization takes place.
```{r, echo=F,include=F}
sd <- openxlsx::read.xlsx(file.path(wd, "sampledescription.xlsx"))
sd[2,1] <- "File_2"
sd[5,1] <- "File_5"
openxlsx::write.xlsx(sd, file.path(wd, "sampledescription.xlsx"), overwrite = T)
```
sampledescription.xlsx:
```{r, echo=F}
knitr::kable(openxlsx::read.xlsx(file.path(wd, "sampledescription.xlsx")))
```
```{r}
fcexpr::sync_sampledescription(FCS.file.folder = file.path(wd, "FCS_files"), write.log = F)
```

The prefixed continuous number is added automatically as well as the .fcs-suffix.  
sampledescription.xlsx:
```{r, echo=F}
knitr::kable(openxlsx::read.xlsx(file.path(wd, "sampledescription.xlsx")))
```
FCS files have been renamed accordingly.
```{r}
list.files(file.path(wd, "FCS_files", "Part_1"))
```
** Note: When files are renamed, respective FlowJo Workspaces have to be opened once in order to adopt the changes in file names in the wsp-file. FlowJo recognizes FCS by other parameters than the file name.**

## Reorder

The FCS files have been added in chronological order of acquisition (see the date in identity column). If one is not happy with that, the row-order in the xlsx-file can be changed. When running sync_sampledescription the prefix-number will be adjusted. **NOTE: Never change the content of the identity column and never mix up rows of FileName and identity!!!**

```{r, echo=F,include=F}
sd <- openxlsx::read.xlsx(file.path(wd, "sampledescription.xlsx"))
sd <- sd[c(1,2,9,10,5,6,3,4,7,8),]
openxlsx::write.xlsx(sd, file.path(wd, "sampledescription.xlsx"), overwrite = T)
```
sampledescription.xlsx:
```{r, echo=F}
knitr::kable(openxlsx::read.xlsx(file.path(wd, "sampledescription.xlsx")))
```
```{r}
fcexpr::sync_sampledescription(FCS.file.folder = file.path(wd, "FCS_files"), write.log = F)
```
```{r}
list.files(file.path(wd, "FCS_files", "Part_1"))
```

## Addition of new FCS files

When FCS files are added to the FCS_files folder and sync_sampledescription is run, the xlsx-file will be appended by the new files (~synchronized).
```{r}
# this command uses example files from the fcexpr package
utils::untar(base::system.file("extdata", "Part_2.tgz", package = "fcexpr"), exdir = file.path(wd, "FCS_files"))
```
```{r}
list.files(file.path(wd, "FCS_files"))
fcexpr::sync_sampledescription(FCS.file.folder = file.path(wd, "FCS_files"), write.log = F)
```
sampledescription.xlsx:
```{r, echo=F}
knitr::kable(openxlsx::read.xlsx(file.path(wd, "sampledescription.xlsx")))
```

## Exclude or remove samples

The function argument exclude.folders in sync_sampledescription specifies folder names that will be ignored for syncing with the xlsx-file. Such folders may contain controls like rainbow beads or compensation controls or similar - in general, files which do not contain biological information. If, by accident, one or more of these files end up in our xlsx-file since one forgot to copy them to a respective folder one can have them removed by deleting the FileName, **but only the FileName**, from the xlsx-file.

```{r, echo=F,include=F}
sd <- openxlsx::read.xlsx(file.path(wd, "sampledescription.xlsx"))
sd$FileName <- ifelse(grepl("comp", sd$FileName), NA, sd$FileName)
openxlsx::write.xlsx(sd, file.path(wd, "sampledescription.xlsx"), overwrite = T)
```
sampledescription.xlsx:
```{r, echo=F}
knitr::kable(openxlsx::read.xlsx(file.path(wd, "sampledescription.xlsx")))
```
```{r}
fcexpr::sync_sampledescription(FCS.file.folder = file.path(wd, "FCS_files"), write.log = F)
```

The prefixed numbers are made continuous by renaming files accordingly.
sampledescription.xlsx:
```{r, echo=F}
knitr::kable(openxlsx::read.xlsx(file.path(wd, "sampledescription.xlsx")))
```

## The log file

By default a log file is written (write.log = T) which documents all changes to sampledescription.xlsx. On Mac/linux it is hidden, on Windows it is not.

```{r, echo=F}
unlink(wd, recursive = T)
```
